NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-03-25 main.tex
Update text

\subsection{Analysing simulations}
In practise, simulation data in the NMRlipids databank can be currently analyzed by first downloading the data using the information in README.yaml and then performing the analysis on local computer utilising the universal naming conventions for molecules and atoms therein. This procedure is illustrated in Fig.~\ref{fig:overlay}D and tables~\ref{tab:codes} and~\ref{tab:codesApplication} list examples of codes that perform such analyses in practise.

\subsection{Calculation of C--H bond order parameters}
The C--H bond order parameters were calculated directly from the carbon and hydrogen positions using the definition
%The order parameters are defined as
\begin{equation}\label{OP}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where angular brackets denote the ensemble average, \textit{i.e.}, average over all sampled configurations of all lipids in a simulation, and $\theta$ is the angle between the C--H bond and the membrane normal. As in previous NMRlipids publications, the order parameters were first calculated separately for each lipid and the standard error of the mean over different lipids was used as the error estimate~\cite{botan15}. The script that calculates C--H bond order parameters from all simulations in the NMRlipids databank is available at \texttt{/Scripts/AnalyzeDatabank/calcOrderParameters.py} in the NMRlipids databank repository. The resulting order parameters are stored for all simulations in files named %\newline 
[\texttt{lipid\_name}]\texttt{OrderParameters.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.

@@

\subsection{Calculation of a bilayer area per lipid and thickness}
Area per lipids of bilayers were calculated by dividing the time-averaged area of the simulation box with the total number of lipid and surfactant molecules (see the list in table~\ref{tab:abbreviations}) in the simulation. The script that calculates area per lipids from all simulations in the NMRlipids databank repository is available at \texttt{Scripts/AnalyzeDatabank/calcAPL.py} in the NMRlipids databank repository. The resulting area per lipids are stored for all simulations in files named \texttt{apl.json} at folders in \texttt{/Data/Simulations}.

Thicknesses of lipid bilayers were calculated from the intersections of lipid and water electron densities. The script that calculates thickness of all simulations in the NMRlipids databank is available at \texttt{Scripts/AnalyzeDatabank/calc\_thickness.py} in the NMRlipids databank repository. The resulting thicknesses are stored in files named \texttt{thickness.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.

\subsection{Principal Components Analysis of equilibration of simulations}
To estimate how well the provided trajectories are converged, the Principal Components Analysis (PCA) was used \cite{buslaev16,buslaev2020principal}. Following the PCALipids \cite{buslaev2020principal} protocol, the lipids of a specific type were first selected, and trajectories of individual lipids were concatenated. Next, each lipid configuration was aligned to the average lipid structure. Afterwards, the PCA was applied on the cartesian coordinates of all heavy atoms of the lipid. Since it was shown before, that the motions along the first, major, principal component (PC) are the slowest ones \cite{buslaev16}, the equilibration of individual lipids was estimated based on the dynamics of the first PC. The suggested way to estimate the characteristic timescales needed to adequately sample the configurational space of individual lipids is to calculate the distribution convergence of the trajectories projected on the first PC\cite{buslaev16}. However, due to the linear dependence between autocorrelation decay times and distribution convergence times, faster calculation times and higher computational stability of autocorrelation decay times\cite{buslaev16, buslaev2020principal}, the distribution convergence timescales were calculated from autocorrelation decay times:
\begin{equation}
    \tau_{\text{convergence}} = k*\tau_{\text{autocorrelation}}
\end{equation}

The empirical coefficient $k$ is equal to $k = 49$ and was calculated based on the analysis of 8 trajectories from the Databank and included simulations of POPC, POPS, POPE, POPG, and DPPC with CHARMM36 force fields. All simulations, used for $k$ calculation, were at least 200 ns long. As was shown previously \cite{buslaev16}, the coefficient does not depend on the force field, thus only CHARMM36 simulations, as the most frequent in the Databank, were used.

As a measure of equilibration quality for each trajectory, the ratio between the convergence time of first PC and the trajectory length, $\tau_{\rm rel} = \tau_{\text{convergence}}/\tau_{\text{sim}}$, was calculated for each lipid type individually. Sterols were excluded from the analysis. The script that calculates the equilibration of lipids is available at \texttt{Scripts/BuildDatabank/NMRPCA\_timerelax.py} in the NMRlipids databank repository. The resulting values are stored in files named \texttt{eq\_times.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.

@@

Also, a quality measure based on differences in simulated and experimental form factors across available $q$-range is defined in the SIMtoEXP program~\cite{kucerka10}. However, the lobe heights in simulated form factors depend on the simulation box size as shown in Fig.~\ref{fig:sizedependence}. Therefore, the quality measure defined in SIMtoEXP  would also depend on the simulation box size. In contrast, locations of form factor minima are independent on simulation box size in Fig.~\ref{fig:sizedependence}. Here we use the location of the first form factor minima for the quality evaluation because automatic detection of the location of second minima is inaccurate for some experimental data due to fluctuations, such as for the POPE data in Figs.~\ref{fig:quality} D and E. The first minimum correlates well with the thickness of a membrane (Fig.~\ref{fig:quality} F), although the correlation of the second minima would be even stronger (Fig.~\ref{fig:QualityCorrelationsSI}). In practise, we first filter the fluctuations from the form factor data using Savitzky--Golay filter (window length 30 and polynomial order 1) and locate the first minima above 0.1\,\AA$^{-1}$ from both simulation and experimental data. The quality of a form factor is then defined as Euclidean distance between the minima in simulated and experimental form factors, $FF_q = |FF_\mathrm{min}^\mathrm{sim}-FF_\mathrm{min}^\mathrm{exp}|$.

>>>>

\subsection{Analysing simulations}
In practise, simulations in the NMRlipids databank can be analyzed by executing a program that loops over README.yaml files in the {\it databank layer}, downloads the data using the information in README.yaml files, and then performs the desired analysis on a local computer utilising the universal naming conventions for molecules and atoms defined in README.yaml and mapping files. This procedure is illustrated in Fig.~\ref{fig:overlay}D and examples of codes that perform such analyses in practise are listed in tables~\ref{tab:codes} and~\ref{tab:codesApplication}.

\subsection{Principal Components Analysis of equilibration of simulations}
To estimate how well conformational ensembles of lipids are converged in trajectories, the Principal Components Analysis (PCA) following the PCALipids protocol was used \cite{buslaev16,buslaev2020principal}. 
%, the lipids of a specific type were first selected, and trajectories of individual lipids were concatenated. 
To this end, each lipid configuration was first aligned to the average structure of that lipid type, and PCA analysis was then applied on the cartesian coordinates of all heavy atoms of the lipid. Because the motions along the first, major, principal component are the slowest ones \cite{buslaev16}, the equilibration of each lipid type was estimated from the ratio between the distribution convergence of the trajectories projected on the first PC and the trajectory length, $\tau_{\rm rel} = \tau_{\text{convergence}}/\tau_{\text{sim}}$~\cite{buslaev16, buslaev2020principal}. If $\tau_{\rm rel} < 1$, simulations can be considered to be sufficiently long for lipid molecules to sample their conformationla ensembles, while in simulations with $\tau_{\rm rel} > 1$ individual molecules may not have fully sample their conformational ensembles. Rigid molecules that do not exhibit significant conformational fluctuations, such as sterols, were excluded from the analysis. 
%The suggested way to estimate the characteristic timescales needed to adequately sample the configurational space of individual lipids is to calculate the distribution convergence of the trajectories projected on the first PC\cite{buslaev16}. 
In practise, the distribution convergence times were calculated utilising its linear dependence on autocorrelation decay times,
%\begin{equation}
    $\tau_{\text{convergence}} = k*\tau_{\text{autocorrelation}}$,
%\end{equation}
because calculation of autocorrelation decay times is faster and computationally more stable than direct calculation of distribution convergence times~\cite{buslaev16, buslaev2020principal}. The empirical coefficient $k$ is equal to $k = 49$ and was calculated based on the analysis of 8 trajectories with the length of more than 200\,ns including simulations of POPC, POPS, POPE, POPG, and DPPC with CHARMM36 force fields. %All simulations, used for $k$ calculation, were at least 200 ns long. 
Because the coefficient does not depend on the force field~\cite{buslaev16}, the value determined from these CHARMM36 simulations ,can be used for all simualtions in the Databank. The script that calculates the equilibration of lipids is available at \texttt{Scripts/BuildDatabank/NMRPCA\_timerelax.py} in the NMRlipids databank repository. The resulting values are stored in files named \texttt{eq\_times.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.

\subsection{Calculation of C--H bond order parameters}
The C--H bond order parameters were calculated directly from the carbon and hydrogen positions using the definition
%The order parameters are defined as
\begin{equation}\label{OP}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where angular brackets denote the ensemble average, \textit{i.e.}, average over all sampled configurations of all lipids in a simulation, and $\theta$ is the angle between the C--H bond and the membrane normal. As in previous NMRlipids publications, the order parameters were first calculated separately for each lipid and the standard error of the mean over different lipids was used as the error estimate~\cite{botan15}. However, order parameters for simulations with $\tau_{\rm rel} > 1$ may be influenced by the starting structure and thereby their error bars may be underestimated. The script that calculates C--H bond order parameters from all simulations in the NMRlipids databank is available at \texttt{/Scripts/AnalyzeDatabank/calcOrderParameters.py} in the NMRlipids databank repository. The resulting order parameters are stored for all simulations in files named %\newline 
[\texttt{lipid\_name}]\texttt{OrderParameters.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.

@@

\subsection{Calculation of a bilayer area per lipid and thickness}
Area per lipids of bilayers were calculated by dividing the time-averaged area of the simulation box with the total number of lipid and surfactant molecules in the simulation (see table~\ref{tab:abbreviations} for the list of molecules considered as lipids or surfactants). The script that calculates area per lipids from all simulations in the NMRlipids databank repository is available at \texttt{Scripts/AnalyzeDatabank/calcAPL.py} in the NMRlipids databank repository. The resulting area per lipids are stored for all simulations in files named \texttt{apl.json} at folders in \texttt{/Data/Simulations}.

Thicknesses of lipid bilayers were calculated from the intersections of lipid and water electron densities. The script that calculates thicknesses of all simulations in the NMRlipids databank is available at \texttt{Scripts/AnalyzeDatabank/calc\_thickness.py} in the NMRlipids databank repository. The resulting thicknesses are stored in files named \texttt{thickness.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.

@@

Also, a quality measure based on differences in simulated and experimental form factors across available $q$-range is defined in the SIMtoEXP program~\cite{kucerka10}. However, the lobe heights in simulated form factors depend on the simulation box size as shown in Fig.~\ref{fig:sizedependence}. Therefore, the quality measure defined in SIMtoEXP  would also depend on the simulation box size. In contrast, locations of form factor minima are independent on simulation box size in Fig.~\ref{fig:sizedependence}. Here we use only the location of the first form factor minimum for the quality evaluation because location of the second minimum is difficult to detect automatically in some experimental data sets due to fluctuations, such as for the POPE data in Figs.~\ref{fig:quality} D and E. The first minimum correlates well with the thickness of a membrane (Fig.~\ref{fig:quality} F), although the correlation of the second minima would be even stronger (Fig.~\ref{fig:QualityCorrelationsSI}). In practise, we first filter the fluctuations from the form factor data using Savitzky--Golay filter (window length 30 and polynomial order 1) and locate the first minima above 0.1\,\AA$^{-1}$ from both simulation and experimental data. The quality of a form factor is then defined as Euclidean distance between the minima in simulated and experimental form factors, $FF_q = |FF_\mathrm{min}^\mathrm{sim}-FF_\mathrm{min}^\mathrm{exp}|$.
